\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-06-12 doc/incertae_sedis.tex
Update.

\caption{\label{fig:Handling-incertae-sedis}Handling \emph{incertae sedis
}taxa recovers additional edges and taxon names.
Supertree construction on input tree (a) and taxonomy tree (b) with \emph{incertae sedis}
clade $C$ leads to synthesis tree (c) if \emph{incertae sedis} annotations are
ignored, but to synthesis tree (d) if \emph{incertae sedis} annotations are respected.
The tree in (d) recovers taxon names $AB$ and $A$, as well as the edge to clade $AB$
that groups $A$ and $B$. Taxon names and edges that are conditional on handling
\emph{incertae sedis} are shaded grey. \mthcomment{We should probably remove $C3$ and replace $B1$ and $B2$ with $B$
to simplify this. We could use triangle to indicate that $D$ and $B$ could be leaves or clades.}
}

@@

The rooted split associated with taxonomy node $n$ serves to determine both when $n$ is in conflict
with an input tree, and to where to place the taxon name for clade $n$ on the synthesis tree.
% Removed mention of incertae sedis possibly affecting subproblem decomposition.  It
% actually doesn't but that is surprising, and therefore kind of a result.
% \paragraph{Definition of rooted splits}
As is typically done, we define rooted splits for each node by noting that
removing the edge connecting the node to its parent would produce two trees.
The leaf label sets of these trees
the partition the full leaf set $\mathcal{L}$ into two groups.
The include set $\mathcal{I}(n)$ which
does not contain the root, and the exclude set $\mathcal{E}(n)$ which does contain the root.
Such a split may be written
\[ \mathcal{I}(n)|\bullet\mathcal{E}(n).\]
where the $\bullet$ indicates the root. 
The include set is the set of labels for leaves that are
descendants of the node.
The exclude set is the set of
labels for leaves that are not descendants of that node.
The ``include set'' and ``exclude set'' of a ``rooted split'' used here are equivalent
    to the ``inside set'' and ``outside set'' of a ``$n$-taxon statement''
    \citep[\emph{sensu}][]{Wilkinson1994}.

@@

\paragraph{Equivalence rooted split definitions and phylogenetic definitions of names}
In terms of phylogenetic nomenclature the supertree pipeline acts as if the
taxonomy defines a conditional node-based name \citep[see][]{deQueiroz2013} for each
higher taxon.
So for the name labeling node $n$ in the taxonomic
tree, the operational definition would be: ``the clade defined by the most recent
common ancestor (MRCA) of $\mathcal{I}(n)$ as long as that MRCA is not an ancestor of
$\mathcal{E}(n)$.''
OTT is not based on a set of explicitly defined phylogenetic name definitions.
In many contexts, a branch-based definition for $n$ (``the clade by all of the taxa more
closely related to all fo them members of $\mathcal{I}(n)$ than to any member of
$\mathcal{E}(n)$, if such a clade exists'') would result in similar behavior
during the tree construction. 
An exception to this (discussed below) arises in the step of assigning names to the 
    supertree.

\paragraph{Equivalence rooted split definitions with character encodings}
Representing a split in a tree by a character is the basis of one of the oldest
    supertree methods: matrix representation with parsimony \citep[MRP, see][]{Baum1992,Ragan1992}.
In the rooted form of such an approach, character state 1 is used for tips that are
    descendants of a internal node in an input tree and 0 is used for other tips.
The partial splits caused by \emph{incertae sedis} annotations naturally map to
    an encoding of missing data (typically denoted with the `?' symbol in a matrix).
Thus the taxonomy shown in Figure \ref{fig:Handling-incertae-sedis}b could be represented as the matrix shown in 
    Figure \ref{fig:Handling-incertae-sedis-mrp}.
In the naive treatment of ignoring the \emph{incertae sedis} annotation, the `?' characters
   in the row for taxa $C1$, $C2$, and $C3$ would be replaced with 0's.

\begin{figure}
\begin{center}
\begin{tabular}{|r|p{2em}|p{2em}|p{2em}|p{2em}|}
\hline
& \multicolumn{4}{|c|}{Character/clade label} \\
leaves & C & AB & B & A \\
\hline
$A1$, $A2$ & 0 & 1 & 0 & 1 \\
$B1$, $B2$ & 0 & 1 & 1 & 0 \\
$C1$, $C2$, $C3$ & 1 & ? & ? & ?  \\
$D$  & 0 & 0 & 0 & 0 \\
\hline
\end{tabular}
\end{center}
\caption{\label{fig:Handling-incertae-sedis-mrp} The matrix encoding 
of an \emph{incertae sedis} interpretation of the taxonomy shown in  Figure \ref{fig:Handling-incertae-sedis}b.
Only the unique character encodings are show, thus the leftmost column
lists all of the tip taxa that would be assigned a particular character encoding.
}

@@

However, there are cases where thus procedure does not result in one name per
node.  This case can occur, for example, when two \emph{incertae sedis} siblings are 
``inter-digitated'', or  combined into a single clade in such a way that their
members cannot be separated.  When the members of the two clades share the same
MRCA, as in Fig. \ref{fig:interdigitation}, then the names for both clades
apply to the same node.  However, in this case, neither name is ancestral to
the other, and the previous solution does not apply.
In such cases, we treat all applicable names as synonyms for the node.
\mthcomment{Need to verify that the namer records this situation as synonyms, as it should!}

@@

One could easily
    imagine a ``pruning decomposer'' that does not allow the say taxon to
    occur in more than one subproblem.
This approach would work in a straightforward way if the phylogenetic inputs were only $\tau_1$ and $\tau_3$,
    yielding the decomposition shown in Figure \ref{fig:decomp}g, but if
    both $\tau_1$ and $\tau_2$ were inputs the decomposer would have to
    resolve conflict about the placement of $\alpha_1$.
Implementing the behavior in the subproblem decomposition step would
    break the separation between the ``divide'' and ``conquer'' steps
    which is a key aspect of propinquity's current efficiency.

@@

Our sub-problem solver naturally handles \emph{incertae sedis} taxa.
This is because we define the semantics of \emph{incertae sedis} taxa
in terms of partial splits, and our solver natively supports building
trees from partial splits through its use of the BUILD algorithm.
Handling \emph{incertae sedis} taxa thus requires loading incertae
sedis information and computing partial splits for \emph{incertae
sedis} taxa before solving a sub-problem.
After solving a sub-problem,
we must apply taxon names from the taxonomy tree to the sub-problem
solution tree.
The solution tree is considered to a fixed tree and not
to have any \emph{incertae sedis} nodes, or any other forms of
uncertainty.

>>>>

\caption{\label{fig:Handling-incertae-sedis}Handling \emph{incertae sedis
}taxa recovers additional edges and taxon names.
Supertree construction on input tree (a) and taxonomy tree (b) with \emph{incertae sedis}
clade $C$ leads to synthesis tree (c) if \emph{incertae sedis} annotations are
ignored, but to synthesis tree (d) if \emph{incertae sedis} annotations are respected.
The tree in (d) recovers taxon names $AB$ and $A$, as well as the edge to clade $AB$
that groups $A$ and $B$. Taxon names and edges that are conditional on handling
\emph{incertae sedis} are shaded grey. \mthcomment{We could use triangle to indicate that $D$ and $B$ could be leaves or clades.}
}

@@

The rooted split associated with taxonomy node $n$ serves to determine both when $n$ is in conflict
with an input tree, and where to place the taxon name for clade $n$ on the synthesis tree.
% Removed mention of incertae sedis possibly affecting subproblem decomposition.  It
% actually doesn't but that is surprising, and therefore kind of a result.
% \paragraph{Definition of rooted splits}
As is typically done, we define rooted splits for each node by noting that
removing the edge connecting the node to its parent would produce two trees.
The leaf label sets of these trees partition the full leaf set $\mathcal{L}$ into two groups: the include
set $\mathcal{I}(n)$ which does not contain the root, and the exclude set $\mathcal{E}(n)$
which does contain the root. Such a split may be written
\[ \mathcal{I}(n)|\bullet\mathcal{E}(n).\]
where the $\bullet$ indicates the root. The include set is the set of labels for leaves that are
descendants of the node. The exclude set is the set of 
labels for leaves that are not descendants of that node.
The ``include set'' and ``exclude set'' of a ``rooted split'' used here are equivalent
    to the ``inside set'' and ``outside set'' of a ``$n$-taxon statement''
    \citep[\emph{sensu}][]{Wilkinson1994}.

@@

\paragraph{Equivalence of rooted split definitions and phylogenetic definitions of names}
In terms of phylogenetic nomenclature the supertree pipeline acts as if the
taxonomy defines a conditional node-based name \citep[see][]{deQueiroz2013} for each
higher taxon.
So for the name labeling node $n$ in the taxonomic
tree, the operational definition would be: ``the clade defined by the most recent
common ancestor (MRCA) of $\mathcal{I}(n)$ as long as that MRCA is not an ancestor of
$\mathcal{E}(n)$.''
OTT is not based on a set of explicitly defined phylogenetic name definitions.
In many contexts, a branch-based definition for $n$ (``the clade by all of the taxa more
closely related to all fo them members of $\mathcal{I}(n)$ than to any member of
$\mathcal{E}(n)$, if such a clade exists'') would result in similar behavior
during the tree construction. 
An exception to this (discussed below) arises in the step of assigning names to the 
    supertree.

\paragraph{Equivalence of rooted split definitions with character encodings}
Representing a split in a tree by a character is the basis of one of the oldest
supertree methods: matrix representation with parsimony \citep[MRP, see][]{Baum1992,Ragan1992}.
In the rooted form of such an approach, character state 1 is used for tips that are
descendants of a internal node in an input tree and 0 is used for other tips and for the root.
The partial splits caused by \emph{incertae sedis} annotations naturally map to
an encoding of missing data (typically denoted with the `?' symbol in a matrix).
Thus the taxonomy shown in Figure \ref{fig:Handling-incertae-sedis}b could be represented as the matrix shown in 
Figure \ref{fig:Handling-incertae-sedis-mrp}.
In the naive treatment of ignoring the \emph{incertae sedis} annotation, the `?' characters
   in the row for taxa $C1$, $C2$, and $C3$ would be replaced with 0's.

\begin{figure}
\begin{center}
\begin{tabular}{|r|p{2em}|p{2em}|p{2em}|p{2em}|}
\hline
& \multicolumn{3}{|c|}{Character/clade label} \\
leaves & C & AB & A \\
\hline
$A1$       & 0 & 1 & 1 \\
$A2$       & 0 & 1 & 1 \\
$B$        & 0 & 1 & 0 \\
$C1$       & 1 & ? & ? \\
$C2$       & 1 & ? & ? \\
  $D$      & 0 & 0 & 0 \\
  root     & 0 & 0 & 0 \\
\hline
\end{tabular}
\end{center}
\caption{\label{fig:Handling-incertae-sedis-mrp} The matrix encoding 
of an \emph{incertae sedis} interpretation of the taxonomy shown in  Figure \ref{fig:Handling-incertae-sedis}b.
The root is coded as a special leaf.
}

@@

However, there are cases where thus procedure does not result in one name per
node.  This case can occur, for example, when two \emph{incertae sedis} siblings are 
``inter-digitated'', or  combined into a single clade in such a way that their
members cannot be separated.  When the members of the two clades share the same
MRCA, as in Fig. \ref{fig:interdigitation}, then the names for both clades
apply to the same node.  However, in this case, neither name is ancestral to
the other, and the previous solution does not apply.
In such cases, we treat all applicable names as synonyms for the node.
\mthcomment{Need to verify that the namer records this situation as synonyms, as it should!}
\bdrcomment{I assume this is not linked to publication}

@@

One could easily
    imagine a ``pruning decomposer'' that does not allow the same taxon to
    occur in more than one subproblem.
This approach would work in a straightforward way if the phylogenetic inputs were only $\tau_1$ and $\tau_3$,
    yielding the decomposition shown in Figure \ref{fig:decomp}g, but if
    both $\tau_1$ and $\tau_2$ were inputs the decomposer would have to
    resolve conflict about the placement of $\alpha_1$.
Implementing the behavior in the subproblem decomposition step would
    break the separation between the ``divide'' and ``conquer'' steps
    which is a key aspect of propinquity's current efficiency.

@@

Our sub-problem solver naturally handles \emph{incertae sedis} taxa.
This is because we define the semantics of \emph{incertae sedis} taxa
in terms of partial splits, and our solver natively supports building
trees from partial splits through its use of the BUILD algorithm.
Handling \emph{incertae sedis} taxa thus requires loading incertae
sedis information and computing partial splits for \emph{incertae
sedis} taxa before solving a sub-problem.
After solving a sub-problem,
we must apply taxon names from the taxonomy tree to the sub-problem
solution tree.
The solution tree is considered to be a fixed tree and not
to have any \emph{incertae sedis} nodes, or any other forms of
uncertainty.
